---
title: "Global Life Expectancy Trends (R)"
engine: knitr
author: "Rohan Chakkadath"
date: 2026-09-23
---

<style>
  /* A light steel blue-grey tint inspired by the official R project logo */
  body {
    background-color: #f0f4f8;
  }
</style>

This post examines historical trends in human longevity and economic development from 1952 to 2007. Through data manipulation and visualization in R, we will track the global median life expectancy over time and compare regional outcomes.

## Dataset Overview

**Data Source:** The [Gapminder dataset](https://www.gapminder.org/data/)  
**License:** Free to use under the [Creative Commons Attribution 4.0 International (CC BY 4.0)](https://creativecommons.org/licenses/by/4.0/) license.

The dataset provides country-level records over 5-year intervals. The `glimpse` output below shows the structure of the data, including population, continent, and life expectancy.

```{r}
#| echo: true
#| warning: false
#| message: false

library(gapminder)
library(dplyr)

glimpse(gapminder)
```

## Continental Overview in 2007

Grouping the 2007 data by continent to calculate median life expectancy and GDP per capita. Using medians reduces the impact of extreme outliers within each continent.

```{r}
#| echo: true
#| code-fold: false

summary_2007 <- gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarise(
    median_life_exp = median(lifeExp),
    median_gdp = median(gdpPercap),
    total_population = sum(as.numeric(pop)),
    num_countries = n()
  ) |>
  arrange(desc(median_life_exp))

knitr::kable(summary_2007, digits = 2, caption = "2007 Continental Summary")
```

## Top 5 Countries by Life Expectancy (2007)

To see which specific nations are driving the upper boundaries of these regional medians, we can isolate the top five countries globally.

```{r}
#| echo: true
#| code-fold: false

top_countries <- gapminder |>
  filter(year == 2007) |>
  arrange(desc(lifeExp)) |>
  select(country, continent, lifeExp, gdpPercap) |>
  head(5)

knitr::kable(top_countries, digits = 2, caption = "Top 5 Countries (2007)")
```

## Longitudinal Life Expectancy Trajectory

Calculating and plotting the global median life expectancy across all available survey years to view the aggregate trend over time.

```{r}
#| echo: true
#| label: fig-life-exp-trend
#| code-fold: false
#| fig-cap: "Global median life expectancy progression from 1952 to 2007."

yearly_trend <- gapminder |>
  group_by(year) |>
  summarise(median_life_exp = median(lifeExp))

plot(
  yearly_trend$year,
  yearly_trend$median_life_exp,
  type = "b",
  pch = 19,
  col = "navy",
  xlab = "Year",
  ylab = "Median Life Expectancy (Years)",
  main = "Global Median Life Expectancy (1952 - 2007)"
)
grid()
```

### Visualizing Continental Life Expectancy (2007)

This bar chart provides a direct visual comparison of median life expectancy across the five continents using the 2007 subset.

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: false

library(ggplot2)

ggplot(summary_2007, aes(x = continent, y = median_life_exp, fill = continent)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Median Life Expectancy by Continent (2007)",
    x = "Continent",
    y = "Median Life Expectancy (years)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")
```

**Observations from 2007 Data:**

*   **Highest Life Expectancy:** Oceania and Europe lead the dataset, with median life expectancies over 76 years.
*   **Middle Range:** The Americas and Asia group closely together in the low 70s.
*   **Lowest Life Expectancy:** Africa has the lowest median life expectancy at 52.9 years, creating a noticeable gap between it and the other four continents.