---
title: "Python and R Interoperability Bonus"
author: "Rohan Chakkadath"
date: 2026-09-23
---

<style>
  /* A soft linear gradient blending R's steel blue and Python's subtle yellow */
  body {
    background: linear-gradient(135deg, #f0f5fa 0%, #fefcf0 100%);
  }
</style>

## Pointing Reticulate to the Virtual Environment

Configuring the `reticulate` package to target the project's isolated `uv` virtual environment guarantees the code executes using project-specific dependencies rather than the global system installation.

```{r}
#| message: false
#| warning: false
library(reticulate)

use_virtualenv("../../.venv", required = TRUE)
```

## Creating Data in R

Initializing a baseline dataframe natively in R.

```{r}
r_data <- data.frame(
  id = 1:5,
  score = c(85, 92, 78, 90, 88)
)
print(r_data)
```

## Reading R Data into Python

Accessing the R environment via the `r.` object in Python. A new column is calculated and appended to the dataframe using `pandas`.

```{python}
import pandas as pd

python_df = r.r_data
python_df['curved_score'] = python_df['score'] + 5

print(python_df)
```

## Returning Python Data back to R

Retrieving the transformed object from the Python environment back into R using the `py$` object.

```{r}
final_r_data <- py$python_df

print(final_r_data)
```