---
title: "Python and R Interoperability Bonus"
author: "Rohan Chakkadath"
date: 2026-09-23
---

<style>
  /* A soft linear gradient blending R's steel blue and Python's subtle yellow */
  body {
    background: linear-gradient(135deg, #f0f5fa 0%, #fefcf0 100%);
  }
</style>

## Pointing Reticulate to the Virtual Environment

Configuring the `reticulate` package to target the project's isolated `uv` virtual environment guarantees the code executes using project-specific dependencies rather than the global system installation.

```{r}
#| message: false
#| warning: false

library(reticulate)
use_virtualenv("../../.venv", required = TRUE)
```

## Creating Data in R

**Data Source:** The [Gapminder dataset](https://www.gapminder.org/data/)  
**License:** Free to use under the [Creative Commons Attribution 4.0 International (CC BY 4.0)](https://creativecommons.org/licenses/by/4.0/) license.

We will load a small sample natively in R to pass into Python.

```{r}
#| message: false
#| warning: false

library(gapminder)

# Grab the first 5 rows of the dataset
r_data <- head(gapminder, 5)
print(r_data)
```

## Reading R Data into Python

Accessing the R environment via the `r.` object in Python. A new column calculating life expectancy in months is appended to the dataframe using `pandas`.

```{python}
import pandas as pd

python_df = r.r_data
python_df['lifeExp_months'] = python_df['lifeExp'] * 12

print(python_df)
```

## Returning Python Data back to R

Retrieving the transformed object from the Python environment back into R using the `py$` object.

```{r}
final_r_data <- py$python_df

print(final_r_data)
```

**Interoperability Highlights:**

*   **Seamless Translation:** The `reticulate` package automatically converts R `data.frame` objects into Python `pandas` DataFrames.
*   **State Retention:** Variables and dataframe columns modified in the Python environment are preserved and instantly accessible when returning to the R workspace.