---
title: "Python and R Interoperability Bonus"
author: "Rohan Chakkadath"
date: 2026-09-23
---

## Pointing Reticulate to the Virtual Environment

By default, R might try to use your system's global Python. To ensure true reproducibility, we explicitly tell the `reticulate` package to use the Python binary inside our `uv` virtual environment.

```{r}
#| message: false
#| warning: false
library(reticulate)

# Force reticulate to use the project's .venv
use_virtualenv("../../.venv", required = TRUE)
```

## Creating Data in R

First, let's create a simple dataframe using R.

```{r}
# Create a small dataset in R
r_data <- data.frame(
  id = 1:5,
  score = c(85, 92, 78, 90, 88)
)
print(r_data)
```

## Reading R Data into Python

Now, we switch to a Python chunk. We can read the R object natively into Python by calling `r.name` (where "name" is the variable we created in R).

```{python}
import pandas as pd

# Access the R dataframe using r.r_data
python_df = r.r_data

# Modify the dataframe using Python (adding a curve to the score)
python_df['curved_score'] = python_df['score'] + 5
print(python_df)
```

## Returning Python Data back to R

Finally, we switch back to an R chunk. We can retrieve the modified dataframe from Python by calling `py$name`.

```{r}
# Access the modified Python dataframe using py$python_df
final_r_data <- py$python_df

# Display the curved scores back in R
print(final_r_data)
```